---
title: "Diagnostic catalog"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostic catalog}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(evalueHMM)
```

This vignette shows the main diagnostic families available in `evalueHMM` on a small simulated trajectory.

```{r data}
set.seed(20260522)

null_parameters <- create_hmm_movement_parameters()
validation_data <- simulate_hmm_movement(
  n_times = 120,
  parameters = null_parameters
)
```

## Full-density diagnostics

A state-number diagnostic compares the null HMM with a diagnostic HMM containing one additional state.

```{r state-diagnostic}
extra_state_parameters <- create_hmm_movement_parameters(
  initial_probs = c(0.55, 0.35, 0.10),
  transition_matrix = matrix(
    c(0.90, 0.08, 0.02,
      0.10, 0.85, 0.05,
      0.12, 0.10, 0.78),
    nrow = 3,
    byrow = TRUE
  ),
  step_shape = c(2, 7, 4),
  step_rate = c(3, 2, 1.5),
  angle_mean = c(0, 0, 0.4),
  angle_sd = c(1.6, 0.35, 0.8)
)

state_diagnostic <- diagnostic_extra_state(
  data = validation_data,
  null_parameters = null_parameters,
  extra_state_parameters = extra_state_parameters
)

summarise_predictive_diagnostic(state_diagnostic)
```

An angular diagnostic keeps the number of states fixed but modifies the angular component.

```{r angle-diagnostic}
angle_parameters <- perturb_hmm_movement_parameters(
  parameters = null_parameters,
  angle_sd_multiplier = c(1.2, 1.4)
)

angle_diagnostic <- diagnostic_angle(
  data = validation_data,
  null_parameters = null_parameters,
  angle_parameters = angle_parameters
)

summarise_predictive_diagnostic(angle_diagnostic)
```

## Feature-level diagnostics

The step-angle diagnostic uses sequential Rosenblatt residuals and a Gaussian copula diagnostic density.

```{r copula-diagnostic}
copula_diagnostic <- diagnostic_step_angle_dependence(
  data = validation_data,
  null_parameters = null_parameters,
  rho = 0.35
)

summarise_predictive_diagnostic(copula_diagnostic)
head(copula_diagnostic$features)
```

## Predictable combinations

Diagnostics can be combined by predictable mixture weights.

```{r mixture}
mixture <- diagnostic_mixture(
  diagnostics = list(
    angle = angle_diagnostic,
    copula = copula_diagnostic
  ),
  weights = c(0.5, 0.5)
)

summarise_predictive_diagnostic(mixture)
```

## Localization

A diagnostic can be localized using predictable weights. The following example only accumulates evidence over a time window.

```{r localization}
window_weights <- time_window_weights(
  time = mixture$path$time,
  start = 40,
  end = 90
)

localized <- localize_predictive_diagnostic(
  diagnostic = mixture,
  weights = window_weights,
  localization_name = "time_window"
)

summarise_predictive_diagnostic(localized)
```

State-localized diagnostics can also use filtered predictive state probabilities.

```{r state-weights}
state_weights <- hmm_predictive_state_weights(
  data = validation_data,
  parameters = null_parameters,
  state = 2
)

summary(state_weights)
```
