---
title: "Getting started with evalueHMM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with evalueHMM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  fig.align = "center",
  dpi = 120
)
```

```{r setup}
library(evalueHMM)
library(ggplot2)
```

`evalueHMM` implements predictive e-diagnostics for hidden Markov models
(HMMs) of animal movement. The core idea is to compare, sequentially, two
observable predictive densities:

1. `p0`, the fitted null HMM used as the reference model;
2. `q`, a diagnostic predictive alternative chosen before evaluating the
   validation observations.

At each time step, the diagnostic increment is

\[
\log E_t - \log E_{t-1} = \log q(Y_t \mid Y_{1:t-1})
  - \log p_0(Y_t \mid Y_{1:t-1}).
\]

Positive increments mean that the diagnostic alternative predicted the next
movement observation better than the null model. The cumulative sum is the
log e-process. A large positive excursion is evidence against the fitted null
generator.

## A small validation experiment

The following example is deliberately small and fully reproducible. We define a
two-state null HMM, then define a diagnostic alternative that has slightly
larger step lengths and less diffuse turning angles. To make the visual signal
easy to see, the validation trajectory is simulated from the diagnostic
alternative. In an applied analysis, the validation data would be held-out data,
not simulated from a known alternative.

```{r simulate}
set.seed(20260522)

null_parameters <- create_hmm_movement_parameters()

diagnostic_parameters <- perturb_hmm_movement_parameters(
  parameters = null_parameters,
  step_mean_multiplier = c(1.25, 1.10),
  angle_sd_multiplier = c(0.75, 0.85)
)

validation_data <- simulate_hmm_movement(
  n_times = 180,
  parameters = diagnostic_parameters,
  n_individuals = 1
)

head(validation_data)
```

The simulated data contain step lengths, turning angles and the latent state
used by the simulator. The latent state is shown below only because this is a
simulation. It is not needed to compute the diagnostic.

```{r trajectory-plot}
make_track <- function(data) {
  heading <- cumsum(data$turning_angle)
  data$x <- cumsum(data$step_length * cos(heading))
  data$y <- cumsum(data$step_length * sin(heading))
  data$state <- factor(data$state)
  data
}

track_data <- make_track(validation_data)

ggplot(track_data, aes(x = x, y = y, colour = state)) +
  geom_path(linewidth = 0.6, alpha = 0.75) +
  geom_point(size = 1.4) +
  coord_equal() +
  scale_colour_manual(values = c("#2A9D8F", "#E9C46A", "#264653")) +
  labs(
    x = "Reconstructed x coordinate",
    y = "Reconstructed y coordinate",
    colour = "Simulated state",
    title = "Validation trajectory used for the diagnostic"
  ) +
  theme_minimal()
```

## Predictive log-densities

The diagnostic only needs the observable predictive log-density under the null
model and under the diagnostic alternative.

```{r predictive-densities}
log_p0 <- hmm_movement_predictive_log_density(
  data = validation_data,
  parameters = null_parameters
)

log_q <- hmm_movement_predictive_log_density(
  data = validation_data,
  parameters = diagnostic_parameters
)

density_comparison <- data.frame(
  time = validation_data$time,
  log_p0 = log_p0,
  log_q = log_q,
  log_e_increment = log_q - log_p0
)

head(density_comparison)
```

The next plot shows the one-step log predictive densities. The two curves need
not be separated at every time step. The e-process accumulates small gains and
losses sequentially.

```{r density-plot}
density_long <- rbind(
  data.frame(time = density_comparison$time, model = "null p0", value = log_p0),
  data.frame(time = density_comparison$time, model = "diagnostic q", value = log_q)
)

ggplot(density_long, aes(x = time, y = value, colour = model)) +
  geom_line(linewidth = 0.65) +
  labs(
    x = "Time",
    y = "One-step log predictive density",
    colour = NULL,
    title = "Observable predictive densities"
  ) +
  theme_minimal()
```

## E-process construction

On the log scale, the e-process is a cumulative sum of the log-density ratios.

```{r eprocess}
eprocess <- compute_eprocess(
  log_p0 = log_p0,
  log_p1 = log_q,
  alpha = 0.05,
  time = validation_data$time
)

diagnostic <- make_predictive_diagnostic(
  diagnostic_name = "step_angle_alternative",
  log_p0 = log_p0,
  log_q = log_q,
  time = validation_data$time
)

summarise_predictive_diagnostic(diagnostic)
```

```{r eprocess-plot}
ggplot(eprocess$path, aes(x = time, y = log_e_cumulative)) +
  geom_hline(
    yintercept = eprocess$threshold,
    linetype = "dashed",
    colour = "#D1495B"
  ) +
  geom_line(colour = "#16324F", linewidth = 0.8) +
  geom_point(size = 1.1, colour = "#16324F") +
  annotate(
    "text",
    x = max(eprocess$path$time),
    y = eprocess$threshold,
    label = "log(1 / alpha)",
    hjust = 1,
    vjust = -0.6,
    colour = "#D1495B"
  ) +
  labs(
    x = "Time",
    y = "Cumulative log e-value",
    title = "Predictive e-process"
  ) +
  theme_minimal()
```

The dashed line is `log(1 / alpha)`. Crossing that line gives evidence against
the null HMM at level `alpha` in the anytime-valid predictive sense used by the
package.

## Interpreting increments

The cumulative path is often easier to interpret together with the individual
increments. Large positive bars are observations for which the diagnostic
alternative predicted better than the null. Negative bars are observations for
which the null predicted better.

```{r increment-plot}
ggplot(eprocess$path, aes(x = time, y = log_e_increment)) +
  geom_col(aes(fill = log_e_increment > 0), width = 0.85, show.legend = FALSE) +
  scale_fill_manual(values = c("#6C757D", "#2A9D8F")) +
  geom_hline(yintercept = 0, colour = "grey35") +
  labs(
    x = "Time",
    y = "Log e-increment",
    title = "Where the evidence accumulates"
  ) +
  theme_minimal()
```

In practice, this diagnostic should be read as model criticism. A signal does
not identify a unique scientific mechanism by itself. It says that the null HMM
is predictively outperformed by the chosen diagnostic alternative on the
validation sequence.
