---
title: "Getting started with evalueHMM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with evalueHMM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

`evalueHMM` implements predictive e-diagnostics for hidden Markov models of movement. The basic workflow is:

1. define a fitted null movement HMM;
2. define a diagnostic predictive alternative before validation;
3. compute observable predictive log-densities under both models;
4. accumulate the log-density ratios as a predictive e-process.

```{r setup}
library(evalueHMM)
```

For a small reproducible example, simulate a trajectory from the built-in two-state movement HMM.

```{r simulate}
set.seed(20260522)

null_parameters <- create_hmm_movement_parameters()
validation_data <- simulate_hmm_movement(
  n_times = 150,
  parameters = null_parameters,
  n_individuals = 1
)

head(validation_data)
```

A diagnostic alternative can be any predictive density fixed before seeing the validation trajectory. Here we perturb the null parameters slightly.

```{r diagnostic-density}
diagnostic_parameters <- perturb_hmm_movement_parameters(
  parameters = null_parameters,
  step_mean_multiplier = c(1.10, 0.95),
  angle_sd_multiplier = c(1.10, 1.20)
)

log_p0 <- hmm_movement_predictive_log_density(
  data = validation_data,
  parameters = null_parameters
)

log_q <- hmm_movement_predictive_log_density(
  data = validation_data,
  parameters = diagnostic_parameters
)
```

The e-process is the cumulative product of the density ratios. On the log scale, this is a cumulative sum of `log_q - log_p0`.

```{r eprocess}
eprocess <- compute_eprocess(
  log_p0 = log_p0,
  log_p1 = log_q,
  alpha = 0.05,
  time = validation_data$time
)

summarise_predictive_diagnostic(
  make_predictive_diagnostic(
    diagnostic_name = "perturbed_null",
    log_p0 = log_p0,
    log_q = log_q,
    time = validation_data$time
  )
)
```

```{r plot}
plot_eprocess(eprocess)
```

The dashed line is `log(1 / alpha)`. Crossing this threshold is evidence against the fitted null generator in the anytime-valid predictive sense described in the manuscript.
