---
title: "Reproducibility guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducibility guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.6,
  fig.align = "center",
  dpi = 120
)
```

```{r setup}
library(ggplot2)
```

This repository has two reproducibility layers:

1. the installable R package `evalueHMM`;
2. the research compendium containing simulations, the elk application and the
   LaTeX manuscript.

The package layer is intentionally small enough for routine checks. The research
layer contains heavier computations and generated outputs used for the
manuscript.

## Workflow overview

The diagram below summarizes the intended workflow. Package checks should pass
without requiring manuscript-scale simulation outputs.

```{r workflow-plot}
workflow <- data.frame(
  x = 1:5,
  y = 1,
  label = c(
    "Source code",
    "Package checks",
    "Vignettes",
    "Research scripts",
    "Manuscript"
  ),
  detail = c(
    "R/, tests/, man/",
    "R CMD check",
    "pkgdown articles",
    "simulations/ and application/",
    "paper/"
  )
)

ggplot(workflow, aes(x = x, y = y)) +
  geom_segment(
    aes(x = 1.15, xend = 4.85, y = 1, yend = 1),
    linewidth = 0.6,
    colour = "grey55"
  ) +
  geom_point(size = 7, colour = "#2A9D8F") +
  geom_text(aes(label = label), nudge_y = 0.16, fontface = "bold", size = 3.3) +
  geom_text(aes(label = detail), nudge_y = -0.16, size = 3.0) +
  scale_x_continuous(limits = c(0.6, 5.4)) +
  scale_y_continuous(limits = c(0.65, 1.35)) +
  labs(title = "Reproducibility layers in the repository") +
  theme_void()
```

## Package checks

From the repository root, the package can be checked quickly without rebuilding
all vignette outputs:

```bash
R CMD build --no-build-vignettes .
R CMD check --no-manual --no-build-vignettes evalueHMM_0.1.0.tar.gz
```

For a full package check with vignette rebuilding:

```bash
R CMD build .
R CMD check --no-manual evalueHMM_0.1.0.tar.gz
```

The continuous integration workflow runs the package check on Linux, macOS and
Windows. The pkgdown workflow builds the online documentation from the same
source files.

```{r check-table}
checks <- data.frame(
  layer = c(
    "Unit tests",
    "Examples",
    "Vignettes",
    "pkgdown site",
    "Research scripts"
  ),
  command = c(
    "testthat::test_local()",
    "R CMD check",
    "R CMD build; R CMD check",
    "pkgdown::build_site()",
    "Rscript simulations/run_all_simulations.R"
  ),
  routine = c(
    "Every package change",
    "Every package change",
    "Before documentation publication",
    "Before GitHub Pages deployment",
    "For manuscript regeneration"
  )
)

checks
```

## Research workflow

The simulation scripts are kept outside the package build because they are
manuscript-scale computations.

```bash
Rscript simulations/run_all_simulations.R
Rscript application/run_application.R
```

The manuscript can be compiled from the `paper/` directory:

```bash
cd paper
pdflatex predictive_e_diagnostics_hmm_improved.tex
pdflatex predictive_e_diagnostics_hmm_improved.tex
pdflatex supplementary_material.tex
pdflatex supplementary_material.tex
```

Generated research outputs are written under:

```text
results/
application/data_processed/
paper/figures/
manuscript_outputs/
```

These outputs are intentionally separated from the source package. That keeps
package installation and checking lightweight while preserving the full research
workflow in the same repository.

## Reproducible environments

The repository includes `renv.lock` for recreating the package development
environment:

```r
renv::restore()
```

The package tests and vignettes use small seeded examples. The manuscript
simulation scripts set their own random seeds and write generated files to the
research-output directories listed above.

## Practical checklist

Before publishing package or manuscript changes, the intended sequence is:

1. run the package tests;
2. rebuild the vignettes and pkgdown site;
3. run `R CMD check`;
4. regenerate manuscript-scale outputs only when scripts or analysis choices
   have changed;
5. compile the manuscript and supplementary material.

This separation makes the package reproducible as software and the manuscript
reproducible as a research compendium.
